---
title: "Simulated Starter Data Analysis Project"
subtitle: ""
author: "Sean O'Sullivan"
date: "2024-06-15"
format:
  html:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis-template-references.bib
csl: ../apa.csl
---

# Findings

**Alex Martinez contributed to this exercise on `2024-06-15`.**

```{r echo=FALSE, include=FALSE}
library(here)
library(dplyr)
library(skimr)
library(ggplot2)
data_location <- here::here("starter-analysis-exercise","data","processed-data","processeddata2.rds")
mydata <- readRDS(data_location)
table_file3 = here::here("starter-analysis-exercise","results", "tables-files", "resulttable3.rds")
linearmodel <- readRDS(table_file3)
```

First we examine our two new variables and how they relate to our prior variables of height and weight.

```{r echo=FALSE}
p5 <- ggplot(mydata, aes(x = Employment, y = Height)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(x = "Employment", y = "Height", title = "Boxplot of Height by Employment") +
  theme_minimal()
plot(p5)
```

When examining boxplots of height across employment status we observe that folks who are employed full-time tend to be taller than those who are employed full-time. However, of note the single unemployed person is among the tallest across all 3 classifications. 

```{r echo=FALSE}
p6 <- mydata[which(!is.na(mydata$Salary)),] %>% 
  ggplot(aes(x = Weight, y = Salary)) +
  geom_point(color = "blue") +
  labs(x = "Weight", y = "Salary", title = "Scatter Plot of Weight vs. Salary") +
  theme_minimal()
plot(p6)
```


When we examine a scatterplot of weight against annual salary in USD we observe that those with middle of the road weights tended to make the most money, with those over- or under-weight earning less.

On the whole it would seem that our simulated data tells the story of pretty privilege, with those that are taller and of well-adjusted weights being more likely to be employed full-time and high-earners.

```{r}
print(linearmodel)
```


Examining a linear model of height as a function of salary and employment status we see predictably no statistical significance for either predictor variables as, in addition to being simulated data, the relationship in question is dubious and nonsensical at best.